---
title: Command lines for microbiota analysis in R
author: Jiayi
date: '2018-10-15'
slug: command-lines-for-microbiota-analysis-in-r
categories:
  - R
tags:
  - microbiota
  - R Markdown
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Inspired by [Riffomonas](http://www.riffomonas.org)

### 0. Before you start  
#### 0.1 Set working directory  
```{r}
setwd(dir = "~/Documents/mothur exercise/rmarkdown/")
```
~ meaning: root

####0.2 Two Sets of Files needed for analysis  
File origin|File type|Notes
------------- | ------------- | ------------- 
Mothur | axes file ended with ".ave.pcoa.axes"|tsv(Tab Separated Values)file  
Mothur | summary file ended with "ave-std.summary"|convert format into xls, csv, or tsv before importing    
Mothur |summary file ended with "tax.summary" | convert format into xls, csv, or tsv before importing  
Metadata|file ended with ".xls", ".xlsx", ".csv", ".tsv", etc...|file(s) containing scription of subject, grouping, treatment, etc...  
  
####0.3 Load libraries needed
```{r , echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(readxl)
```

###1. Import files into R  
* For excel file(s):  
  1. use command "read excel"  
  2. define the data type for each column using argument "col_types"  
eg:  

metadata <- read_excel("ex3 design.xlsx", 
                       col_types = c("text", "text", "text", 
                                     "text", "skip", "skip", "skip", 
                                     "skip"))


to be continued...

