---
title: code for microbiota analysis in R
author: Jiayi
date: '2018-10-15'
slug: code-for-microbiota-analysis-in-r
categories:
  - R
tags:
  - microbiota
  - R Markdown
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Inspired by [Riffomonas](http://www.riffomonas.org)

###0. Before you start  
####0.1 Set working directory  
```{r}
setwd(dir = "~/Documents/mothur exercise/rmarkdown/")
```

####0.2 Two Sets of Files needed for analysis  
File origin|File type|Notes
------------- | ------------- | ------------- 
Mothur | axes file ended with ".ave.pcoa.axes"|tsv(Tab Separated Values)file  
Mothur | summary file ended with "ave-std.summary"|convert format into xls, csv, or tsv before importing    
Mothur |summary file ended with "tax.summary" | convert format into xls, csv, or tsv before importing  
Metadata|file ended with ".xls", ".xlsx", ".csv", ".tsv", etc...|file(s) containing scription of subject, grouping, treatment, etc...  
  
####0.3 Load libraries needed
```{r , echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(readxl)
```
  
###1. Import files into R  
* For excel file(s):  
  1. use command "read excel"  
  2. define the data type for each column using argument "col_types"  
eg: 
```{r xlsx, message=FALSE, warning=FALSE, , echo=TRUE, paged.print=FALSE}
metadata <- read_excel("ex3 design.xlsx", 
                       col_types = c("text", "text", "text", 
                                     "text", "skip", "skip", "skip", 
                                     "skip"))
```
* For ".axe"" and ".summary" (which are basicly csv/tsv) file(s):  
  1. use command "read_csv"/"read_tsv",  
  2. define data type for "group" as "character"
eg:
```{r t s v}
pcoa <- read_tsv(file="stability.opti_mcc.thetayc.0.03.lt.ave.pcoa.axes",
                 col_types=cols(group=col_character()
                 )
)
alpha <- read_tsv(file="report.opti_mcc.0.03.subsample.groups.ave-std.summary",
                  col_types=cols(group = col_character())) %>%
  filter(method=='ave') %>%
  select(group, sobs, shannon, invsimpson, coverage)
```
* For taxonomy files:   
    + option 1: use excel to trim data and import to R afterwards
    + option 2: transcode the original version (in csv/tsv/xls format) and trim in R(more hereof later)
```{r taxonomy}
tax <- read_excel("tax.xlsx")
```

###2. Trim taxonomy data and innerjoin dataframes  
####2.1 Trim & get clean data
```{r trim & get clean data, warning=FALSE}
#be careful do not change row or column order (which would be confusing after innerjoin with metadata framse)
taxlevel <- as.numeric(tax[1,]) #subtract taxlevel row
genus_ <- tax[, which(taxlevel=="6")] ##subset genus columns (taxlevel = 6)
genus_ <-cbind(tax[ ,1], genus_) #bind 1st column as rownames reference

genus_2 <- genus_[-c(1, 4, 5), ] #remove rows: taxlevel, daughterleves, total
genus_colomname <- paste(genus_2[2, ], genus_2[1, ]) ##creat a vector containing column names, use paste function to concatenate both genus name and rank ID
genus_colomname[1] <- "taxon" #substitute the 1st element with "taxon" ==> 1st column name will subsequently be "taxon" instead of previous pasted version. 

genus_3 <- genus_2[3:23, ]
colnames(genus_3) <- genus_colomname #set up colomn names using the vector "genus_colomname"" generated before??????

genus_4 <- genus_3[ , -c(1)] #discard sample information from genus_3 (important: do not reorder rows/columns)
genus_4[] <- lapply(genus_4, function(x) as.numeric(as.character(x))) #change type into ???numeric???

genus_4 <- as.matrix(genus_4) #change into a matrix (array) so as to use "prop.table" function afterwards
genus_5 <- prop.table(genus_4, 1) #calculate x/sum(x) by row, 2 then by column
genus_6 <- 100*genus_5 #show in percentage

genus_6 <- as.data.frame(genus_6)

countif <- numeric(0)
for (i in 1:ncol(genus_6)){
  countif[i] <- length(which(genus_6[[i]] >= 1))#count in column: how many cells >= 1 -->return with boolean-->sum up
  countif
  genus_7 <- rbind(countif , genus_6)
}


zero <- which(genus_7[1,] == 0) #remove columns where abundance are all <1
genus_8 <- genus_7[-1 , -zero]

genus <- cbind(genus_3[ ,1], genus_8)
colnames(genus)[1] <- "taxon"

rm(genus_, genus_2, genus_3, genus_4, genus_5, genus_6, genus_7, genus_8, tax, countif, genus_colomname, genusname, i, taxlevel, zero) #remove proc files (which are unnecessary for subsequent analysis)
```
  
####2.2 Calculating differences between baseline and afterward timepoints 
```{r 2.2 Do the subtraction}

```
  
  
####2.2 Innerjoin dataframes  
```{r warning=FALSE}
metadata_pcoa <- inner_join(metadata, pcoa, by = c("sample" = "group")) 
metadata_genus <- inner_join(metadata, genus, by = c("sample" = "taxon"))
metadata_alpha <- inner_join(metadata, alpha, by = c("sample" = "group"))
metadata_pcoa_genus <- inner_join(metadata_pcoa, genus, by = c("sample" = "taxon"))
```

###3. Barplot  
```{r Barplot, message=FALSE}
##building a data frame contains: group with salmonella#
genus_by_fixative <- metadata_genus %>%
    group_by(fixative) %>%
    summarize(mean_genus = mean(genus$`Bifidobacterium 0.1.1.1.2.1.3`), sd_genus = sd(genus$`Bifidobacterium 0.1.1.1.2.1.3`))
  
  #plot for genus#
  ggplot(genus_by_fixative, aes(x=fixative, y=mean_genus, fill=fixative)) +
    geom_errorbar(aes(width=0.5, ymax=mean_genus+sd_genus, ymin=mean_genus-0.1*mean_genus)) +
    geom_col(show.legend=FALSE)+
    scale_fill_manual(name=NULL, 
                      values=c("darkgreen","orange"),
                      breaks=c("0", "1"),
                      labels=c("0", "1") ) + 
    scale_x_discrete(limits=c("0","1"),
                     labels=c("0","1")) +
    scale_y_continuous(labels = scales::percent) +
    labs(title="Relative Abundance of genus",
         x=NULL,
         y="Abundance")
```
  
###4. Scatter plot  
4.1 Basic Version  
```{r Basic version}
ggplot(data=metadata_pcoa_genus, mapping=aes(x=axis1, y=axis2, color=fixative, label=`time days`)) +
  geom_point(shape=19, size= 2) + 
  coord_fixed(ratio = 1) + #The ratio represents the number of units on the y-axis equivalent to one unit on the x-axis.
  labs(title = "PCoA by affecte genera abundance", 
       x = "PCoA Axis1", 
       y = "PCoA Axis2") +
  geom_text(aes(label= paste(metadata_pcoa_genus$subject, metadata_pcoa_genus$`time days`, metadata_pcoa_genus$fixative, sep = "/")),hjust=0, vjust=-0)
```  
  
4.2 Upgraded Version  
Set size of each dot as genus abundance of its correspondent sample: 
```{r Upgraded version}
ggplot(data=metadata_pcoa_genus, mapping=aes(x=axis1, y=axis2, color=fixative, label=`time days`)) +
  geom_point(shape=19, size= metadata_pcoa_genus$`Bifidobacterium 0.1.1.1.2.1.3`) + 
  coord_fixed(ratio = 1) + #The ratio represents the number of units on the y-axis equivalent to one unit on the x-axis.
  labs(title = "PCoA by affecte genera abundance", 
       x = "PCoA Axis1", 
       y = "PCoA Axis2") +
  geom_text(aes(label= paste(metadata_pcoa_genus$subject, metadata_pcoa_genus$`time days`, metadata_pcoa_genus$fixative, sep = "/")),hjust=0, vjust=0)
``` 


###5. Wilcoxon test
```{r Wilcoxon, warning=FALSE}
group_metadata_genus <- cbind(groupname = paste(metadata$subject, metadata$fixative, sep = ""), metadata_genus)
for (i in 6:10){
  wilcox.test <- pairwise.wilcox.test(g=group_metadata_genus[["groupname"]], x=group_metadata_genus[[i]], p.adjust.method="BH")
  print(wilcox.test[["p.value"]])
}

```